Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques

### 2019-10-07 paper/paperV2.tex
JFP Proof

Scenarios generated by \textit{simScene} should cover most real-life situations while remaining perceptually plausible. To guarantee the realism of created sound scenes, the information needed for \textit{simScene} generation is based on a corpus of 74 recordings from the GRAFIC project~\cite{aumond2017}. These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, and range from 55~s to 4.5~m in duration. In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}). The available annotations include:

@@

An listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as to obtain subjective annotations for the design and evaluation of the time of presence estimation model. Given the aim of this test, a separate corpus is constructed. This corpus contains 100 sound scenes, including 75 simulated scenes as well as 6 recorded and 19 replicated scenes from the GRAFIC project.

@@

In order to validate the generation procedure of new scenarios, the listening test corpus is extended with replications of recorded sound scenes. Reference recordings are obtained from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations (noted P1-19) with diverse environments. For each of the 19 corresponding recordings, 45 seconds of audio in a single channel are extracted. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The manual annotations available in~\cite{gloaguen2017} in terms of background and event information are then used to replicate the sound scenes using \textit{simScene}. Furthermore, the original 45~s recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization. The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.\\

During the test, participants are asked to evaluate sound scenes in terms of 8 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, tranquil} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. Additionally, to assess the perceived source activity 3 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Time of presence of traffic, voice and bird sources: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

@@

Next, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied to validate the use of simulated sound scenes with new scenarios as well as reduced source complexity. It is obtained by performing a principal components analysis on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects are projected onto the principal components space for each scene and represented in Figure~\ref{fig:pspace_sim} as dots for simulated scenes corpus and crosses for recorded and replicated scenes. These projections show that the space covered by scenes based on new scenarios covers that of the studied real-life environments. This further demonstrates the effectiveness of the scene generation procedure in terms of diversity of scenarios.

@@

\begin{figure}[th]
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/pca_p1.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25). Arrows show discrepancies between corresponding recorded and replicated scenes. For the P1 recording location the standard deviation of the projection of individual responses is shown using ellipses.}\label{fig:pspace_rec}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/pca_sim.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75). Assessments of simulated scenes are projected as dots, and recorded and replicated scenes are projected as crosses.}\label{fig:pspace_sim}
\end{figure}

@@

To train and evaluate a deep learning architecture for sound source recognition using the corpus described in Section~\ref{sec:data_corp}, the activity of sound sources present in each scene must be known. Though, the manual annotation of such information is impractical: it is extremely time-consuming due to the size of the corpus, and the process must be repeated for each studied sound source. Several indicators are identified in the literature that that correlate well with perceptual parameters~\cite{aumond2017, gontier2018, ricciardi2014}. These indicators may be useful to automatically annotate the deep learning corpus without the need for additional human input.

@@

Multilinear regression models of pleasantness are built as a function of the overall loudness and the source-specific time of presence of traffic, voices and birds, with assessments averaged for each scene (n=100). All 15 combinations of the 4 predictors are considered, and to ensure that no multi-collinearity is present between predictors a variance inflation factor (VIF) test is performed prior to each regression. Only combinations for which all predictors verify $VIF<5$ are considered valid. The best resulting model in terms of $R^2_{adj}$ with statistically significant coefficient estimates ($p<0.05$) is:

@@

Though, the performance of the trained model and its capacity to generalize to recorded data rely on several conditions. First, the perceptual characteristics of the simulated corpus should be comparable to that of typical urban environments. According to the results of the listening test presented in Section~\ref{sec:data_val}, the simulation process of sound scenes does not significantly affect the relations between abstract or content-related subjective descriptors, even with newly generated scenarios and a reduced source taxonomy. This is further demonstrated by the model of pleasantness found in Section~\ref{sec:app_mdls} from subjective annotations of source activity in this experiment. The perceptual models on the listening test corpus are similar to those found in previous studies with both \textit{in situ} questionnaires~\cite{aumond2017} and laboratory experiments using recordings~\cite{ricciardi2014}. Second, the quality of predictions is bound by that of annotations. Here, the proposed indicator does not perfectly correspond with the perceived time of presence obtained during the listening test. Furthermore, the indicator is computed using two parameters optimized on available subjective data, though it was found to be quite robust by using a cross-validation scheme during its optimization. While the quality of predictions from the model are encouraging, testing on a larger corpus is required to fully assess its generalization capabilities.

The methodology proposed in this paper can be extended to include additional sound sources provided that enough excerpts of isolated source occurrences are available. This is because the scene simulation process, the time of presence indicator, and the deep learning model are all independent from the considered taxonomy. Of course, the complexity of the learning process scales with the number of considered simultaneously active sources. Particularly, differentiating between sources with similar spectral shapes may yield higher error rates as the deep learning architecture relies on the identification of patterns in third octave spectra. To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching~\cite{salamon2017-2, lasseck2018}.

>>>>

Scenarios generated by \textit{simScene} should cover most real-life situations while remaining perceptually plausible. To guarantee the realism of created sound scenes, the information needed for \textit{simScene} generation is based on a corpus of 74 recordings from the GRAFIC project~\cite{aumond2017}. These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, and range from 55~s to 4.5~m in duration. In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}). The available annotations include:

@@

A listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as to obtain subjective annotations for the design and evaluation of the time of presence estimation model. Given the aim of this test, a separate corpus is constructed. This corpus contains 100 sound scenes, including 75 simulated scenes as well as 6 recorded and 19 replicated scenes from the GRAFIC project.

@@

In order to validate the generation procedure of new scenarios, the listening test corpus is extended with replications of recorded sound scenes. Reference recordings are obtained from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations (noted P1-19) with diverse environments. For each of the 19 corresponding recordings, 45 seconds of audio in a single channel are extracted. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The manual annotations available in~\cite{gloaguen2017} of background and event information are then used to replicate the sound scenes using \textit{simScene}. Furthermore, the original 45~s recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization. The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.\\

During the test, participants are asked to evaluate sound scenes using 8 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, tranquil} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. Additionally, to assess the perceived source activity 3 questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Time of presence of traffic, voice and bird sources: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

@@

Next, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied to validate the use of simulated sound scenes with new scenarios as well as reduced source complexity. It is obtained by performing a principal components analysis on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects for active individuals (simulated scenes) are projected onto the principal components space for and represented in Figure~\ref{fig:pspace_sim} as dots. The assessments for recorded and replicated scenes are then projected as supplementary individuals on this space and represented as crosses. These projections show that the space covered by scenes based on new scenarios covers that of the studied real-life environments. This further demonstrates the diversity of scenarios created by the scene generation procedure.

@@

\begin{figure}[th]
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/pca_p1.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25). Arrows show discrepancies between corresponding recorded and replicated scenes. For the P1 recording location the standard deviation of the projection of individual responses is shown using ellipses.}\label{fig:pspace_rec}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/pca_sim.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75). Assessments of simulated scenes (active individuals) are projected as dots, and recorded and replicated scenes (supplementary individuals) are projected as crosses.}\label{fig:pspace_sim}
\end{figure}

@@

To train and evaluate a deep learning architecture for sound source recognition using the corpus described in Section~\ref{sec:data_corp}, the activity of sound sources present in each scene must be known. Though, the manual annotation of such information is impractical: it is extremely time-consuming due to the size of the corpus, and the process must be repeated for each studied sound source. Several indicators are identified in the literature that correlate well with perceptual parameters~\cite{aumond2017, gontier2018, ricciardi2014}. These indicators may be useful to automatically annotate the deep learning corpus without the need for additional human input.

@@

Multilinear regression models of pleasantness are built as a function of the overall loudness and the source-specific time of presence of traffic, voices and birds, with assessments averaged for each scene (n=100). All possible combinations of variables among the 4 predictors (15 combinations) are considered, and to ensure that no multi-collinearity exists between predictors, a variance inflation factor (VIF) test is performed prior to each regression. Only combinations for which all predictors verify $VIF<5$ are considered valid. The best resulting model in terms of $R^2_{adj}$ with statistically significant coefficient estimates ($p<0.05$) is:

@@

Though, the performance of the trained model and its capacity to generalize to recorded data rely on several conditions. First, the perceptual characteristics of the simulated corpus should be comparable to that of typical urban environments. According to the results of the listening test presented in Section~\ref{sec:data_val}, the simulation process of sound scenes does not significantly affect the relations between abstract or content-related subjective descriptors, even with newly generated scenarios and a reduced source taxonomy. This is further demonstrated by the model of pleasantness found in Section~\ref{sec:app_mdls} from subjective annotations of source activity in this experiment. The perceptual models on the listening test corpus are similar to those found in previous studies with both \textit{in situ} questionnaires~\cite{aumond2017} and laboratory experiments using recordings~\cite{ricciardi2014}. Second, the quality of predictions is bounded by that of annotations. Here, the proposed indicator does not perfectly correspond with the perceived time of presence obtained during the listening test. Furthermore, the indicator is computed using two parameters optimized on available subjective data, though it was found to be quite robust by using a cross-validation scheme during its optimization. While the quality of predictions from the model are encouraging, testing on a larger corpus is required to fully assess its generalization capabilities.

The methodology proposed in this paper can be extended to include the identification of additional sound sources, though it requires the availablity of isolated samples to simulate enough diverse sound scenes to train the detection architecture. This is because the scene simulation process, the time of presence indicator, and the deep learning model are all independent from the considered taxonomy. Of course, the complexity of the learning process scales with the number of considered simultaneously active sources. Particularly, differentiating between sources with similar spectral shapes may yield higher error rates as the deep learning architecture relies on the identification of patterns in third octave spectra. To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching~\cite{salamon2017-2, lasseck2018}.
